---
format:
  html:
    footnotes-hover: false
    css: styles.css
    toc: true
    toc-title: On this page
    toc-location: left
mainfont: "GT-America"
linkcolor: "#00355a"
execute:
  warning: false
  message: false
---


## Variables to Explore

These are based on the Reentry 2030 Dashboard Analysis Plan: https://csgorg.sharepoint.com/:w:/r/sites/Team-JC-Research/Shared%20Documents/CRD_Reentry%202030/Deliverables/Analysis%20Plan%20for%20Reentry%202030%20Dashboard.docx?d=wb7166caab9de4394a552b401c6bcee08&csf=1&web=1&e=A6M2up


**Recidivism:**

* *inc_to_inc_recid_y1*: Incarceration to incarceration recidivism within one year

* *inc_to_inc_recid_y2*: Incarceration to incarceration recidivism within two years

* *inc_to_inc_recid_y3*: Incarceration to incarceration recidivism within three years

* *inc_to_inc_recid_y4*: Incarceration to incarceration recidivism within four years

* *inc_to_inc_recid_y5*: Incarceration to incarceration recidivism within five years

<br>

**Employment:**

*Poverty*

* *inc_above_poverty_y1*: Proportion of individuals above poverty, one year after release from prison

* *inc_above_poverty_y3*: Proportion of individuals above poverty, three years after release from
prison

* *inc_above_poverty_y5*: Proportion of individuals above poverty, five years after release from
prison

*W2*

* *inc_any_w2_y1*: Proportion of individuals with at least one W2 filed, one year after release from
prison

* *inc_any_w2_y3*: Proportion of individuals with at least one W2 filed, three years after release from
prison

* *inc_any_w2_y5*: Proportion of individuals with at least one W2 filed, five years after release from
prison

*Wages*

* *inc_w2_wages_y1*: Income of individuals, one year after release from prison

* *inc_w2_wages_y3*: Income of individuals, three years after release from prison

* *inc_w2_wages_y5*: Income of individuals, five years after release from prison

<br>

**Housing**

* *inc_hud_y1*: Proportion of individuals that received HUD rental housing assistance, one year after release from prison

* *inc_hud_y3*: Proportion of individuals that received HUD rental housing assistance, three years after release from prison

* *inc_hud_y5*: Proportion of individuals that received HUD rental housing assistance, five years after release from prison

<br>

**Healthcare**

*SSI*

* *inc_ssi_y1*: Proportion of individuals who received SSI, one year after release from prison

* *inc_ssi_y3*: Proportion of individuals who received SSI, three years after release from prison

* *inc_ssi_y5*: Proportion of individuals who received SSI, five years after release from prison

*Medicaid*

* *inc_medicaid_y1*: Proportion of individuals enrolled in Medicaid, one year after release from prison

* *inc_medicaid_y3*: Proportion of individuals enrolled in Medicaid, three years after release from
prison

* *inc_medicaid_y5*: Proportion of individuals enrolled in Medicaid, five years after release from
prison

*Medicare*

* *inc_medicare_y1*: Proportion of individuals enrolled in Medicare, one year after release from prison

* *inc_medicare_y3*: Proportion of individuals enrolled in Medicare, three years after release from
prison

* *inc_medicare_y5*: Proportion of individuals enrolled in Medicare, five years after release from
prison

*Mortality*

* *inc_mortality_y1*: Proportion of individuals who have passed away, one year after release from
prison

* *inc_mortality_y3*: Proportion of individuals who have passed away, three years after release
from prison

* *inc_mortality_y5*: Proportion of individuals who have passed away, five years after release from


## Call libraries and set paths

```{r}
# ### install version of highcharter that is compatible with our accessibility requirements 
# remotes::install_github("jbkunst/highcharter", ref = remotes::github_pull(780))

library(tidyverse)
library(scales)
library(csgjcr)
library(highcharter)
library(htmlwidgets)
library(reactable)
library(reactablefmtr)
library(sf)
library(janitor)
library(tidycensus)

```

```{r}
# ### set project path (only need to run once)
# csg_set_project_path(project = "reentry_2030_cjars",
#                      sp_folder = "C:/Users/abyrum/OneDrive/The Council of State Governments/JC Research - Documents/CRD_Reentry 2030",
#                      force = TRUE)

### pull overall reentry 2030 folder path -- that we set in above script
reentry_2030_cjars <- csg_get_project_path("reentry_2030_cjars")

### set path for importing raw cjars data
reentry_2030_cjars_raw <- file.path(reentry_2030_cjars,
                                    "Background",
                                    "CJARS Documentation")


### set path for exporting clean cjars data
reentry_2030_cjars_clean <- file.path(reentry_2030_cjars,
                                      "Data",
                                      "CJARS",
                                      "Clean")

### create list of reentry states to include in analysis
### based on 6 states listed here: https://reentry2030.org/
### NOTE: Missouri and Alabama appear to be missing from file (29 and 01)
### so I've excluded them from the list 
reentry_states_to_include <- c("36", "53", "37", "31")
```

<br>

## Clean state-level data

```{r }
### import state_data.csv
### file note: confirm that file is aggregated by fips, cohort_year, sex, race, age_group, off_type, and repeat_contact
df_cjars_q3_state_raw <- readr::read_csv(file.path(reentry_2030_cjars_raw,
                                            "2023Q3_state_delim", 
                                            "state_delim",
                                            "state_data.csv")) |> 
  ### add state name using fips code
  mutate(state_name = csgjcr::csg_state_convert(fips,
                                                from = "fips",
                                                to = "name"),
         state_abbr = csgjcr::csg_state_convert(fips,
                                                from = "fips",
                                                to = "abbr"),
         ### create flag for reentry states/non-reentry states 
         reentry_state_flag = ifelse(fips%in%reentry_states_to_include,
                                     "Reentry 2030 States","Non-Reentry 2030 States")) |>
  ### reorder state_name after fips
  ### and just keep variables of interest
  ### including one sample size var per data series
  dplyr::select(fips, 
                state_name, 
                state_abbr,
                reentry_state_flag,
                cohort_year:repeat_contact,
                N_inc_to_inc_recid_y1,
                inc_to_inc_recid_y1,
                inc_to_inc_recid_y2,
                inc_to_inc_recid_y3,
                inc_to_inc_recid_y4,
                inc_to_inc_recid_y5,
                N_inc_above_poverty_y1,
                inc_above_poverty_y1,
                inc_above_poverty_y3,
                inc_above_poverty_y5,
                N_inc_any_w2_y1,
                inc_any_w2_y1,
                inc_any_w2_y3,
                inc_any_w2_y5,
                N_inc_w2_wages_y1,
                inc_w2_wages_y1,
                inc_w2_wages_y3,
                inc_w2_wages_y5,
                N_inc_hud_y1,
                inc_hud_y1,
                inc_hud_y3,
                inc_hud_y5,
                N_inc_ssi_y1,
                inc_ssi_y1,
                inc_ssi_y3,
                inc_ssi_y5, 
                N_inc_medicaid_y1,
                inc_medicaid_y1,
                inc_medicaid_y3, 
                inc_medicaid_y5, 
                inc_medicare_y1, 
                inc_medicare_y3, 
                inc_medicare_y5, 
                N_inc_mortality_y1,
                inc_mortality_y1, 
                inc_mortality_y3, 
                inc_mortality_y5) |> 
  ### now clean following grouping vars: sex, race, age_group, off_type, and repeat_contact
  ### 0 == overall counts for all grouping vars will be NA in clean variable versions
  
  ### age
  ### use factor function to assign labels to values
   mutate(age_group_clean = factor(age_group,
         levels = c(1,2,3),
         labels = c("15-24", 
                    "25-40", 
                    "40+")),  
  ### race/ethnicity
  race_clean = factor(race,
         levels = c(1,2,3,4,5),
         labels = c("White, non-Hispanic",
                    "Black, non-Hispanic",
                    "Asian or Pacific Islander, non-Hispanic",
                    "Hispanic",
                    "American Indian or Alaska Native")),
  ### sex
  sex_clean = factor(sex,
         levels = c(1,2),
         labels = c("Male", 
                    "Female")),
  ### off_type
  off_type_clean = factor(off_type,
         levels = c(1,2,3),
         labels = c("Violent", 
                    "Property", 
                    "Drug")),
  ### repeat_contact
  repeat_contact_clean = factor(repeat_contact,
         levels = c(1,2,3),
         labels = c("First-time contact only", 
                    "Repeat contact only",
                    "All calculated with reweights"))) 


### create analytic file with only overall counts and rates
df_cjars_q3_state_overall_only <- df_cjars_q3_state_raw |> 
  ### 0 == overall counts for all grouping vars
  filter(age_group==0,
         race==0, 
         sex==0,
         off_type==0,
         repeat_contact==0) |> 
  ### drop clean grouping vars
  dplyr::select(-c(age_group_clean:repeat_contact_clean))

```

<br>

## Clean county-level data

**Note:** This is just for NC for now -- will need to write a function to loop cleaning over every state and then export rds file. This works if we only use county-level file is for map and table on state page (and don't need comparisons across states)

```{r }
########################################
### pull in county fips codes and names
### for join to county CJARS file
#######################################

### county-level file
### from ACS file
us_county_to_join <- get_acs(
  geography = "county",
  variables = "S0101_C01_001",
  year = 2022,
  geometry = TRUE,
  keep_geo_vars = TRUE,
  cb = TRUE,
  resolution = "20m") |>
  clean_names() |>
  mutate(state_geoid = substr(geoid, 1, 2)) |>
  ### now drop geometry so that we can use as regular df (and not a spatial object)
  ### we initially needed geometry in order to extract county name without label (e.g., county, parish, area, etc.)
  st_drop_geometry() |> 
  ### finally, keep and rename columns we'll use later
  dplyr::select(county_fips = geoid,
                county_name = name_x) |>
  distinct() 

```


```{r }
### county file is large -- run gc first
gc()

### import county_data.csv via data.table::fread b/c of file size
### file note: confirm that file is aggregated by fips, cohort_year, sex, race, age_group, off_type, and repeat_contact
df_cjars_q3_county_raw <- data.table::fread(file.path(reentry_2030_cjars_raw,
                                                      "2023Q3_county_delim", 
                                                      "county_delim",
                                                      "county_data.csv"),
                                            verbose=TRUE) |> 
  ### add leading zeros to fips county code in order to make code 5 digits 
  ### (when leading zeros are dropped, the code is only four digits -- e.g., Arizona)
  mutate(county_fips = str_pad(fips, 5, pad = "0"),
         ### now extract first 2 digits as state fips code
         state_fips = substr(county_fips,1,2)) |> 
  ### TEMP: FILTER ON NC ONLY FOR NOW
  filter(state_fips==37) |> 
  ### join in county name via fips code (pulling from census data)
  left_join(us_county_to_join,
            by = "county_fips") |> 
  ### add state name using fips code
  mutate(state_name = csgjcr::csg_state_convert(state_fips,
                                                from = "fips",
                                                to = "name"),
         state_abbr = csgjcr::csg_state_convert(state_fips,
                                                from = "fips",
                                                to = "abbr"),
         ### create flag for reentry states/non-reentry states 
         reentry_state_flag = ifelse(state_fips%in%reentry_states_to_include,
                                     "Reentry 2030 States","Non-Reentry 2030 States")) |>
  ### reorder state_name after fips
  ### and just keep variables of interest
  ### including one sample size var per data series
  dplyr::select(county_fips, 
                county_name,
                state_fips,
                state_name, 
                state_abbr,
                reentry_state_flag,
                cohort_year:repeat_contact,
                N_inc_to_inc_recid_y1,
                inc_to_inc_recid_y1,
                inc_to_inc_recid_y2,
                inc_to_inc_recid_y3,
                inc_to_inc_recid_y4,
                inc_to_inc_recid_y5,
                N_inc_above_poverty_y1,
                inc_above_poverty_y1,
                inc_above_poverty_y3,
                inc_above_poverty_y5,
                N_inc_any_w2_y1,
                inc_any_w2_y1,
                inc_any_w2_y3,
                inc_any_w2_y5,
                N_inc_w2_wages_y1,
                inc_w2_wages_y1,
                inc_w2_wages_y3,
                inc_w2_wages_y5,
                N_inc_hud_y1,
                inc_hud_y1,
                inc_hud_y3,
                inc_hud_y5,
                N_inc_ssi_y1,
                inc_ssi_y1,
                inc_ssi_y3,
                inc_ssi_y5, 
                N_inc_medicaid_y1,
                inc_medicaid_y1,
                inc_medicaid_y3, 
                inc_medicaid_y5, 
                inc_medicare_y1, 
                inc_medicare_y3, 
                inc_medicare_y5, 
                N_inc_mortality_y1,
                inc_mortality_y1, 
                inc_mortality_y3, 
                inc_mortality_y5) |> 
  ### now clean following grouping vars: sex, race, age_group, off_type, and repeat_contact
  ### 0 == overall counts for all grouping vars will be NA in clean variable versions
  
  ### age
  ### use factor function to assign labels to values
   mutate(age_group_clean = factor(age_group,
         levels = c(1,2,3),
         labels = c("15-24", 
                    "25-40", 
                    "40+")),  
  ### race/ethnicity
  race_clean = factor(race,
         levels = c(1,2,3,4,5),
         labels = c("White, non-Hispanic",
                    "Black, non-Hispanic",
                    "Asian or Pacific Islander, non-Hispanic",
                    "Hispanic",
                    "American Indian or Alaska Native")),
  ### sex
  sex_clean = factor(sex,
         levels = c(1,2),
         labels = c("Male", 
                    "Female")),
  ### off_type
  off_type_clean = factor(off_type,
         levels = c(1,2,3),
         labels = c("Violent", 
                    "Property", 
                    "Drug")),
  ### repeat_contact
  repeat_contact_clean = factor(repeat_contact,
         levels = c(1,2,3),
         labels = c("First-time contact only", 
                    "Repeat contact only",
                    "All calculated with reweights"))) 


### create analytic file with only overall counts and rates
df_cjars_q3_county_overall_only <- df_cjars_q3_county_raw |> 
  ### 0 == overall counts for all grouping vars
  filter(age_group==0,
         race==0, 
         sex==0,
         off_type==0,
         repeat_contact==0) |> 
  ### drop clean grouping vars
  dplyr::select(-c(age_group_clean:repeat_contact_clean))

```


<br>

### Export Files

```{r }
### export both state- and county-level files for dashboard viz

#################
### state level
#################

### overall only
write_rds(df_cjars_q3_state_overall_only,
          file.path(reentry_2030_cjars_clean,"cjars_clean_state_overall_only.rds"))


### with disaggregations
write_rds(df_cjars_q3_state_raw,
          file.path(reentry_2030_cjars_clean,"cjars_clean_state_with_breakdowns.rds"))


#################
### county level
#################

### overall only
write_rds(df_cjars_q3_county_overall_only,
          file.path(reentry_2030_cjars_clean,"cjars_clean_county_overall_only.rds"))


### with disaggregations
write_rds(df_cjars_q3_county_raw,
          file.path(reentry_2030_cjars_clean,"cjars_clean_county_with_breakdowns.rds"))


```


